Nitric oxide feedback to ciliary photoreceptor cells gates a UV avoidance circuit

### 2023-06-07 Manuscript/Manuscript.Rmd
Ms: Methods edit, key resources table update

---
title: Nitric oxide feedback to ciliary photoreceptor cells gates a UV avoidance circuit
author: 
  - name: "Kei Jokura<sup>1,`*`</sup>, Nobuo Ueda<sup>2</sup>, Martin Gühmann<sup>3</sup>, Luis Alfonso Yañez-Guerra<sup>1</sup>, Piotr Słowiński<sup>1</sup>, Kyle C. A. Wedgewood<sup>1</sup> and Gáspár Jékely<sup>1,4,`*`</sup>"
    affiliation: "<sup>1</sup>Living Systems Institute, University of Exeter, Stocker road, Exeter, EX4 4QD, United Kingdom <sup>2</sup>Okinawa Institute of Science and Technology, Okinawa, Japan, <sup>3</sup>School of Biological Sciences, University of Bristol, Bristol, United Kingdom, <sup>4</sup>Centre for Organismal Studies (COS), University of Heidelberg, 69120 Heidelberg, Germany  <br> <sup>*</sup>Correspondence`:` k.jokura @ exeter.ac.uk, gaspar.jekely @ cos.uni-heidelberg.de"
output:
  html_document:
    theme: lumen
  pdf_document:
    latex_engine: lualatex
    template: default
  word_document: default
geometry: left=0.8in,right=0.8in,top=0.8in,bottom=0.5in
mainfont: Arial
monofont: Times New Roman
documentclass: article
classoption: onecolumn
linestretch: 1.2
fontsize: 10pt
bibliography: references.bib
csl: elife.csl
editor_options: 
  markdown: 
    wrap: sentence
---

@@

```{r echo=FALSE}
library(readxl)
key_resource_tale <- read_excel("key_resource_tale/key_resource_tale.xlsx")
df_key_resource_tale <- as.data.frame(key_resource_tale)
knitr::kable(df_key_resource_tale, format = "html", booktabs = TRUE)
```

@@

Before designing the small guide RNA (sgRNA) for the sgRNA:Cas9 nuclease, splice sites and polymorphic sites in our laboratory culture were identified to avoid them.
The sgRNA targeted the third exon of Platynereis dumerilii NOS (target site: 5'-GGGCAATACTGGCTCCACTC-3').
The sgRNA was assembled from two annealed oligonucleotides (5'-TAGGGCAATACTGGCTCCACTC-3', 5'-AAACGAGTGGAGCCAGTATTGC-3') forming overhangs for cloning into a BsaI site of the plasmid pDR27456 (Hwang et al. 2013)(42250, Addgene), which contains next to the BsaI site a tracrRNA sequence.
The plasmid was then used to PCR amplify DNA (primers: T7, 5'-AAAAGCACCGACTCGGTGCC-3') for synthesizing the sgRNA.
The DNA was purified with the QIAquick PCR Purification Kit (Qiagen).
From the DNA, the sgRNA was synthesized with the MEGAshortscript Kit (Thermo Fisher Scientific) and was purified with the MEGAclear Kit (Thermo Fisher Scientific).
Cas9-mRNA was transcribed, capped, and polyA-tailed with the mMessage mMachine Kit and the Poly(A) Tailing Kit (both Thermo Fisher Scientific) from a plasmid (pUC57-T7-RPP2-Cas9) containing the Cas9 ORF fused to 169 base pair 5' UTR from the Platynereis dumerilii 60S acidic ribosomal protein P2.
The sgRNA (18 ng/ml) and the Cas9-mRNA (180 ng/µl) were coinjected into fertilized eggs of Platynereis dumerilii wild-type parents according to an established injection procedure [@Conzelmann2013a].
The eggs were kept at 18°C for 45 min before injection and were injected at 14.5°C.
The injected individuals were kept at 18°C for 5 to 8 days in 6-well- plates (Nunc multidish no. 150239, Thermo Scientific) and then cultured at 22°C until sexual maturity.
The mature worms were crossed to wild-type worms and the progeny was genotyped, resulting in two founder lines, which were bred to homozygosity.

**NOS sequencing and genotyping**

@@

**Vertical column setup for measuring photoresponses**

Photoresponses of larvae of different ages were assayed in a vertical Plexiglas column (31 mm x 10 mm x 160 mm water height).
The column was illuminated from top with light from a monochromator (Polychrome II, Till Photonics).
The monochromator was controlled by AxioVision 4.8.2.0 (Carl Zeiss MicroImaging GmbH) via analog voltage.
The light passed a collimator lens (LAG-65.0-53.0-C with MgF2 Coating, CVI Melles Griot) before entering the column.
The column was illuminated from both sides with light-emitting diodes (LEDs).
The LEDs on each side were grouped into two strips.
One strip contained UV (395 nm) LEDs (SMB1W-395, Roithner Lasertechnik) and the other infrared (810 nm) LEDs (SMB1W-810NR-I, Roithner Lasertechnik).
The UV LEDs were run at 4 V to stimulate the larvae in the column from the side.
The infrared LEDs were run at 8 V (overvoltage) to illuminate the larvae for the camera (DMK 22BUC03, The Imaging Source), which recorded videos at 15 frames per second and was controlled by IC Capture (The Imaging Source).

**Comparing behavior of wildtype and NOS-knockout 3-day-old larvae**

To compare the behavior of wildtype and NOS-knockout larvae at 3 days in the vertical column, the larvae were mixed and left in the dark for 5 min.
The larvae were treated with NOS inhibitors for pharmacology.
The NOS inhibitors were L-NAME.
The larvae were treated with different concentrations in adjacent columns.
The concentrations for the NOS inhibitors were control, 1 mM, 0.1 mM.
The larvae were recorded for 1 min in the dark followed by exposure to collimated cyan (480 nm) light from the top of the column for 2 min, then 2 min darkness, and finally collimated UV (395 nm) light from the top of the column for 2 min.
Stimulus light was provided by the monochromator (Polychrome II, Till Photonics).
Scripts are available at <https://github.com/JekelyLab/NOS>.

**NOS Identification and Phylogenetic Analysis**

To identify NOS, we obtained a "seed" database of oxygenase domain in Pfam database, PF02898.
From these sequences, we produced a Hidden Markov Model (HMM) and used this to mine the 47 metazoan species, 2 choanoflagellate species and 2 filasterea species investigated.
HMM models were run in HMMR3 with an e-value of 1e−15.
We ran CD-Hit [@Fu2012] to eliminate redundant sequences (at a 80% threshold).
We aligned the sequences with MAFFT version 7, with the iterative refinement method E-INS-i.
Alignments were trimmed with TrimAl in gappy-out mode [@Capella-Gutiérrez2009].
To calculate maximum-likelihood trees, we used IQ-tree2 with the LG+G4 model.
To calculate branch support, we ran 1,000 replicates with the aLRT-SH-like and aBayes methods [@Minh2020].
The sequences used for the phylogenetic analysis are available in **Supplementary File 1**, the trimmed alignment is available in **Supplementary File 2** and the pre-trimmed data in **Supplementary file 3**.

**NIT-GC Identification and Phylogenetic Analysis**

To identify NIT-GCs, we obtained a "seed" database of Adenylate and Guanylate cyclase catalytic domain in Pfam database, PF00211.
From these sequences, we produced a Hidden Markov Model (HMM) and used this to mine the 45 metazoan species, 2 choanoflagellate species and 2 filasterea species investigated.
HMM models were run in HMMR3 with an e-value of 1e−15.
We ran CD-Hit [@Fu2012] to eliminate redundant sequences (at a 80% threshold).
To identify clusters, we used the convex-clustering option with 100 jack-knife replicates.
The NIT-GCs are extremely well conserved in membrane-bound guanylate cyclases and form an easily recognizable cluster.
To analyze the phylogeny of NIT-GCs, the cluster containing these GCs together with membrane-bound guanylate cyclases were parsed and used for tree building.
We aligned the sequences with MAFFT version 7, with the iterative refinement method E-INS-i.
Alignments were trimmed with TrimAl in gappy-out mode [@Capella-Gutiérrez2009].
To calculate maximum-likelihood trees, we used IQ-tree2 with the LG+G4 model.
To calculate branch support, we ran 1,000 replicates with the aLRT-SH-like and aBayes methods [@Minh2020].
The sequences used for the phylogenetic analysis are available in **Supplementary File 1**, the trimmed alignment is available in **Supplementary File 2** and the pre-trimmed data in **Supplementary file 3**.

@@

We used Achim et al. for the single-cell data [@Achim2015].
In Williams et al., they used 107 cells as neurons by removing duplicates from Achim et al. single-cell data, so we used those cells [@williams2017].
Since the raw data were read count data, we normalized them to TPM using Python.
After that we converted them to log10.
From the sum of the expression levels in 107 cells for each gene, We calculated the percentage in each cell.
For each cell, we identified them with marker genes.
After created the data in Python, plotted it using R dot plots.
RPKM calculates the total number of reads per million bp and then divides by the length of each gene, so it is not possible to compare between samples.
Instead, TPM first divides by the length of the gene and then divides by the total number of reads per million bases, which allows for more accurate comparisons between samples.
In this case we wanted to compare between samples, so we used TPM.
The total TPM of each gene between the samples was used to calculate the percentage of expressed genes.
The total TPM values for each gene and the percentage of expressed genes are available in **Supplementary file 1**.

@@

Larvae were fixed and treated with Proteinase K, according to the conventional WMISH protocol [@Tessmar-Raible2005], with fixation in 4% paraformaldehyde/ PTW (PBS with 0.05% Tween20) for 2 hr at room temperature, and Proteinase K treatment in 100 µg/ml Proteinase K/ PTW for 3 min [@Tessmar-Raible2005].
Specifically, for the HCR protocol, samples were processed in 1.5 ml tubes.
Probe hybridization buffer, probe wash buffer, amplification buffer, and fluorescent HCR hairpins were purchased from Molecular Instruments (Los Angeles, USA).
Hairpins associated with the b2 initiator sequence were labeled with Alexa Fluor 647, and the hairpins associated with the b3 initiator sequence were labeled with Alexa Fluor 546.
To design probes for HCR, we used custom software [@Kuehn2022] to create 20 DNA oligo probe pairs specific to P. dumerilii NOS, NIT-GC1, NIT-GC2, RYa-pNP (GenBank accession: JF811330.1), and MLD/pedal 2-pNP (GenBank accession: KF515945.1).
The NOS, NIT-GC1 and NIT-GC2 probes were designed to be associated with the b2 initiator sequence, while the RYa-pNP and MLD/pedal 2-pNP probes were designed to be associated with the b3 initiator sequence.
For the detection stage, samples were pre-hybridized in 200 µl of probe hybridization buffer for 1 hr at 37°C, and then incubated in 250 µl hybridization buffer containing probe oligos (4 pmol/ml) overnight at 37°C.
To remove excess probe, samples were washed 4× with 1 ml hybridization wash buffer for 15 min at 37°C, and subsequently 2× in 1 ml 5× SSCT (5× SSC with 0.1% Tween20) for 5 min at room temperature.
For the amplification stage, samples were pre-incubated with 100 µl of amplification buffer for 30 min, room temperature, and then incubated with 150 µl amplification buffer containing fluorescently labeled hairpins (40nM concentration (2ul of 3uM stock in 150ul amplification buffer, snap-cooled as described; [@Choi2018]) overnight in the dark at 25°C.
To remove excess hairpins, samples were washed in 1 ml 5× SSCT at room temperature, twice for 5 min, twice for 30 min, and once for 5 min.
During the first 30 min wash, samples were counterstained with DAPI (Cat. #40043, Biotium, USA).

@@

Whole-mount immunostaining of 2 day old Platynereis larvae fixed with 4% paraformaldehyde were carried out using primary antibodies raised against NIT-GC1, NIT-GC2, NOS, RYamide neuropeptide, RGWamide neuropeptide in rabbit, plus a commercial antibody raised against acetylated tubulin in mouse (Sigma T7451).
The synthetic peptides contained an N-terminal Cys that was used for coupling during purification.
Antibodies were affinity purified from sera as previously described [@Conzelmann2012].
Immunostainings were carried out as previously described [@Conzelmann2012].
The NOS promoter (fragment sizes: 12 Kb) was amplified and cloned upstream of 3xHA- Palmi-tdTomato.
Larvae injected with promoter constructs (ca. 250 ng/ml) were analysed for reporter expression at 3 days post fertilization using an AxioImager Z.1 fluorescence wide-field microscope (Carl Zeiss GmbH, Jena) and immediately fixed for immunostainings.
The protocol followed for immunostaining of HA-tagged reporters was recently described [@veraszto2017].
Specimens were imaged with a LSM 780 NLO or LSM 880 with Airysan Confocal Microscope (Zeiss, Jena).

@@

For calcium imaging, 49--55 hpf larvae were used.
Experiments were performed at room temperature and larvae were immobilised by being embedded in 2.5% agarose filtered artificial seawater between a slide and coverslip spaced with adhesive tape.
GCaMP6s mRNA (1 mg/ml) was injected into zygotes as described previously [@Randel2014a].
Larvae were imaged on a Zeiss LSM 880 with Airyscan (with a C-Apochromat 63X/1.2 Corr - water) with a frame rate of 1.88 frame/sec and an image size of 512 x 512 pixels.
The larvae were stimulated in a region of interest (a circle with 50 pixel diameter) with 405 nm lasers controlled by the Bleaching mode.
The imaging laser had a similar intensity than the stimulus laser but covered an area that was 10 times larger than the stimulus ROI.

@@

Green cGull was used for the cGMP assay [@Matsuda2017].
A full-length Pdum-NIT-GC1 and -NIT-GC2 coding sequences were amplified by PCR starting from a *Platynereis dumerilii* cDNA library and cloned into the pcDNA3.1(+) vector using the T2A self-cleaving sequence.
Cos-7 cells with low expression of endogenous soluble guanylate cyclase were used as cultured cells for gene expression.
This cell line was purchased from Angio-proteomie (CAT no. cAP-0203).
The Cos-7 cells were maintained at 37 °C in 35mm dishes (Nunc™ Glass Bottom Dishes) containing 3 mL of DMEM, high glucose glutamax medium (Thermo; Cat. No. 10566016) supplemented with 10% fetal bovine serum (Thermo; Cat. No. 10082147).
Upon reaching confluency of approximately 85%, we transfected the cells with the plasmid containing Green cGull-T2A-NITGC1.
Transfections were carried out with 150 ng of each plasmid and 0.3 μl of the transfection Lipofectamine 3000 Reagent (invitrogen; Cat. No. L3000001).
Two days post-transfection, we removed the culture medium and substituted it for fresh DMEM-medium.
For single-wavelength imaging experiments, cells in 35-mm dishes were washed twice and imaged in modified Ringer's buffer (140 mM NaCl, 3.5 mM KCl, 0.5mM NaH2PO4 , 0.5mM S-3 MgSO4 , 1.5 mM CaCl2 , 10 mM HEPES, 2 mM NaHCO3 and 5 mM glucose).
Dishes were mounted on a stage heated at 37 °C and imaging was performed using an inverted microscope (LSM880, Zeiss) equipped with an oil-immersion objective lens (UApo/340, 40, NA = 0.17).
Images were acquired using a xenon lamp, 460--495 nm excitation filter, 505-nm dichroic mirror and 510-- 550-nm emission filter (Zeiss).
S-Nitroso-N-acetyl-D, L-penicillamine (SNAP) was purchased from Sigma-Aldrich (St. Louis, MO, USA) .
The exposure time of the EM-CCD camera was controlled by the ZEN software (Zeiss).
Images were acquired every 15 s for 10 min and stimulation was initiated 2 min after starting image acquisition.
Imaging data analysis was performed using ImageJ (National Institutes of Health, Bethesda, MD, USA).

>>>>

---
title: Nitric oxide feedback to ciliary photoreceptor cells gates a UV avoidance circuit
author: 
  - name: "Kei Jokura<sup>1,`*`</sup>, Nobuo Ueda<sup>2</sup>, Martin Gühmann<sup>3</sup>, Luis Alfonso Yañez-Guerra<sup>1</sup>, Piotr Słowiński<sup>1</sup>, Kyle C. A. Wedgewood<sup>1</sup> and Gáspár Jékely<sup>1,4,`*`</sup>"
    affiliation: "<sup>1</sup>Living Systems Institute, University of Exeter, Stocker road, Exeter, EX4 4QD, United Kingdom <sup>2</sup>Okinawa Institute of Science and Technology, Okinawa, Japan, <sup>3</sup>School of Biological Sciences, University of Bristol, Bristol, United Kingdom, <sup>4</sup>Centre for Organismal Studies (COS), University of Heidelberg, 69120 Heidelberg, Germany  <br> <sup>*</sup>Correspondence`:` k.jokura @ exeter.ac.uk, gaspar.jekely @ cos.uni-heidelberg.de"
output:
  html_document:
    theme: lumen
  word_document: default
  pdf_document:
    latex_engine: lualatex
    template: default
geometry: left=0.8in,right=0.8in,top=0.8in,bottom=0.5in
mainfont: Arial
monofont: Times New Roman
documentclass: article
classoption: onecolumn
linestretch: 1.2
fontsize: 10pt
bibliography: references.bib
csl: elife.csl
editor_options: 
  markdown: 
    wrap: sentence
---

@@

```{r echo=FALSE}
library(readxl)
key_resource_tale <- read_excel("key_resource_tale/key_resource_tale.xls")
df_key_resource_tale <- as.data.frame(key_resource_tale)
knitr::kable(df_key_resource_tale, format = "html", booktabs = TRUE)
```

@@

For the generation of a *NOS* knockout we used an sgRNA targeting the third exon of the *Platynereis dumerilii* *NOS* (target site: 5'-GGGCAATACTGGCTCCACTC-3').
We selected the site to avoid polymorphic sites in our laboratory culture.
The sgRNA was assembled from two annealed oligonucleotides (5'-TAGGGCAATACTGGCTCCACTC-3', 5'-AAACGAGTGGAGCCAGTATTGC-3') forming overhangs for cloning into the BsaI site of the plasmid pDR27456 [\@hwang2013](42250,%20Addgene), which contains next to the BsaI site a tracrRNA sequence.
We used this plasmid to PCR amplify DNA (primers: T7, 5'-AAAAGCACCGACTCGGTGCC-3') for synthesizing the sgRNA.
To purify the PCR product, we used the QIAquick PCR Purification Kit (Qiagen).
We then synthesized sgRNA with the MEGAshortscript Kit (Thermo Fisher Scientific) and purified it with the MEGAclear Kit (Thermo Fisher Scientific).
To produce the Cas9-mRNA, we used the plasmid (pUC57-T7-RPP2-Cas9) containing the Cas9 ORF fused to 169 base pair 5' UTR from the *Platynereis dumerilii* 60S acidic ribosomal protein P2.
After in vitro transcription with the mMessage mMachine Kit (Thermo Fisher Scientific), we capped and polyA-tailed the Cas9-mRNA with the Poly(A) Tailing Kit (Thermo Fisher Scientific).
We coinjected the sgRNA (18 ng/ml) and the Cas9-mRNA (180 ng/µl) into fertilized eggs of *Platynereis dumerilii* wild-type parents according to an established injection procedure [@Conzelmann2013a].
The eggs were kept at 18°C for 45 min before injection and were injected at 14.5°C.
The injected individuals were kept at 18°C for 5 to 8 days in 6-well-plates (Nunc multidish no. 150239, Thermo Scientific) and then cultured at 22°C until sexual maturity.
The mature worms were crossed to wild-type worms and the progeny was genotyped, resulting in two founder lines, which were bred to homozygosity.

**Genotyping of NOS alleles**

@@

**Vertical column setup to measure photoresponses**

We assayed larval photoresponses in a vertical Plexiglas column (31 mm x 10 mm x 160 mm water height).
For stimulation, we illuminated the column from the top with a monochromator (Polychrome II, Till Photonics) controlled by AxioVision 4.8.2.0 (Carl Zeiss MicroImaging GmbH) via analog voltage.
The light passed a collimator lens (LAG-65.0-53.0-C with MgF2 Coating, CVI Melles Griot) before entering the column.
The column was illuminated from both sides with light-emitting diodes (LEDs).
The LEDs on each side were grouped into two strips.
One strip contained UV (395 nm) LEDs (SMB1W-395, Roithner Lasertechnik) and the other infrared (810 nm) LEDs (SMB1W-810NR-I, Roithner Lasertechnik).
The UV LEDs were run at 4 V. The infrared LEDs were run at 8 V (overvoltage) to illuminate the larvae for imaging with a DMK camera (DMK 22BUC03, The Imaging Source).
We recorded videos at 15 frames per second with the software IC Capture (The Imaging Source).

To compare the behavior of wildtype and NOS-knockout larvae or larvae treated with L-NAME, we first mixed the larvae and left them in the dark for 5 min.
We then imaged larvae for 1 min in the dark (only IR LED for illumination) followed by exposure to non-directional UV light from both sides, then 2 min darkness, then collimated cyan (480 nm) light from the top of the column for 2 min, then 2 min darkness, and finally collimated UV (395 nm) light from the top of the column for 2 min.
Scripts are available at <https://github.com/JekelyLab/NOS>.

**Identification and Phylogenetic Analysis of NOS sequences**

To identify NOS sequences, we obtained a seed database of oxygenase domains from the Pfam database (PF02898).
From these sequences, we produced a Hidden Markov Model (HMM) and used this to mine the 47 metazoan, 2 choanoflagellate and 2 filasterea species investigated.
HMM models were run in HMMR3 with an e-value of 1e−15.
We ran CD-Hit [@Fu2012] to eliminate redundant sequences (at a 80% threshold).
We aligned the sequences with MAFFT version 7, with the iterative refinement method E-INS-i.
Alignments were trimmed with TrimAl in gappy-out mode [@Capella-Gutiérrez2009].
To calculate maximum-likelihood trees, we used IQ-tree2 with the LG+G4 model.
To calculate branch support, we ran 1000 replicates with the aLRT-SH-like and aBayes methods [@Minh2020].
The sequences used for the phylogenetic analysis are available in **Supplementary File 1**, the trimmed alignment is available in **Supplementary File 2** and the pre-trimmed data in **Supplementary file 3**.

**Identification and Phylogenetic Analysis of NIT-GC sequences**

To identify NIT-GCs, we obtained a seed database of Adenylate and Guanylate cyclase catalytic domains from the Pfam database (PF00211).
From these sequences, we produced a Hidden Markov Model (HMM) and used this to mine the 45 metazoan, 2 choanoflagellate and 2 filasterea species investigated.
HMM models were run in HMMR3 with an e-value of 1e−15.
We ran CD-Hit [@Fu2012] to eliminate redundant sequences (at a 80% threshold).
To identify clusters, we used sequence-similarity-based clustering in Clans [@frickey2004] and the convex-clustering option with 100 jack-knife replicates.
The NIT-GCs are extremely well conserved within the membrane-bound guanylate cyclases and form an easily recognizable cluster.
To analyze the phylogeny of NIT-GCs, the cluster containing these GCs together with membrane-bound guanylate cyclases were parsed and used for tree building.
We aligned the sequences with MAFFT version 7, with the iterative refinement method E-INS-i.
Alignments were trimmed with TrimAl in gappy-out mode [@Capella-Gutiérrez2009].
To calculate maximum-likelihood trees, we used IQ-tree2 with the LG+G4 model.
To calculate branch support, we ran 1000 replicates with the aLRT-SH-like and aBayes methods [@Minh2020].
The sequences used for the phylogenetic analysis are available in **Supplementary File 4**, the trimmed alignment is available in **Supplementary File 5** and the pre-trimmed data in **Supplementary file 6**.

@@

We used a single-cell RNA sequencing dataset from *Platynereis* larvae [@Achim2015] that was further filtered to 107 cells [@williams2017].
We normalized raw read-count data to TPM and converted to log10.
To calculate % expression, we used the sum of the expression levels in the 107 cells for each gene.
The total TPM of each gene between the samples was used to calculate the percentage of expressed genes.
We processed the data in Python and plotted in R.
Individual cells were identified by suits of well-characterised marker genes and spatial mapping [@Achim2015].
The total TPM values for each gene and the percentage of expressed genes are available in **Supplementary file 1**.

@@

Larvae were fixed and treated with Proteinase K, according to the conventional WMISH protocol [@Tessmar-Raible2005], with fixation in 4% paraformaldehyde/ PTW (PBS with 0.05% Tween20) for 2 hr at room temperature, and Proteinase K treatment in 100 µg/ml Proteinase K/ PTW for 3 min [@Tessmar-Raible2005].
Specifically, for the HCR protocol, samples were processed in 1.5 ml tubes.
Probe hybridization buffer, probe wash buffer, amplification buffer, and fluorescent HCR hairpins were purchased from Molecular Instruments (Los Angeles, USA).
Hairpins associated with the b2 initiator sequence were labeled with Alexa Fluor 647, and the hairpins associated with the b3 initiator sequence were labeled with Alexa Fluor 546.
To design probes for HCR, we used custom software [@Kuehn2022] to create 20 DNA oligo probe pairs specific to *P. dumerilii* *NOS*, *NIT-GC1*, *NIT-GC2*, *RYa proneuropeptide* (GenBank accession: JF811330.1), and *MLD/pedal peptide 2 proneuropeptide* (GenBank accession: KF515945.1).
The *NOS*, *NIT-GC1* and *NIT-GC2* probes were designed to be associated with the b2 initiator sequence, while the *RYa* and *MLD/pedal peptide 2* probes were designed to be associated with the b3 initiator sequence.
For the detection stage, samples were pre-hybridized in 200 µl of probe hybridization buffer for 1 hr at 37°C, and then incubated in 250 µl hybridization buffer containing probe oligos (4 pmol/ml) overnight at 37°C.
To remove excess probe, samples were washed 4× with 1 ml hybridization wash buffer for 15 min at 37°C, and subsequently 2× in 1 ml 5× SSCT (5× SSC with 0.1% Tween20) for 5 min at room temperature.
For the amplification stage, samples were pre-incubated with 100 µl of amplification buffer for 30 min, room temperature, and then incubated with 150 µl amplification buffer containing fluorescently labeled hairpins (40nM concentration (2ul of 3uM stock in 150ul amplification buffer, snap-cooled as described; [@Choi2018]) overnight in the dark at 25°C.
To remove excess hairpins, samples were washed in 1 ml 5× SSCT at room temperature, twice for 5 min, twice for 30 min, and once for 5 min.
During the first 30 min wash, samples were counterstained with DAPI (Cat. #40043, Biotium, USA).

@@

Whole-mount immunostaining of *Platynereis* larvae fixed with 4% paraformaldehyde was carried out using the primary antibodies listed in KEY RESOURCES TABLE. Immunostainings were carried out as previously described [@Conzelmann2012].

**Antibody generation and purification**

The synthetic peptides contained an N-terminal Cys that was used for coupling during purification.
Antibodies were affinity purified from sera as previously described [@Conzelmann2012].

**Transient transgenesis**

The *NOS* promoter (fragment sizes: 12 Kb) was amplified and cloned upstream of 3xHA- Palmi-tdTomato.
Larvae injected with promoter constructs (ca. 250 ng/ml) were analysed for reporter expression at 3 days post fertilization using an AxioImager Z.1 fluorescence wide-field microscope (Carl Zeiss GmbH, Jena) and immediately fixed for immunostainings.
The protocol followed for immunostaining of HA-tagged reporters was recently described [@veraszto2017].
Specimens were imaged with a LSM 780 NLO or LSM 880 with Airysan Confocal Microscope (Zeiss, Jena).

@@

For calcium imaging, 49-55 hpf larvae were used.
Experiments were carried out at room temperature and larvae were immobilised by being embedded in 2.5% agarose filtered artificial seawater between a slide and coverslip spaced with adhesive tape.
GCaMP6s mRNA (1 mg/ml) was injected into zygotes as described previously [@Randel2014a].
Larvae were imaged on a Zeiss LSM 880 with Airyscan (with a C-Apochromat 63X/1.2 Corr - water) with a frame rate of 1.88 frame/sec and an image size of 512 x 512 pixels.
The larvae were stimulated in a region of interest (a circle with 50 pixel diameter) with 405 nm lasers controlled by the Bleaching mode.
The imaging laser had a similar intensity than the stimulus laser but covered an area that was 10 times larger than the stimulus ROI.

@@

Green cGull was used for the cGMP assay [@Matsuda2017].
A full-length Pdum-NIT-GC1 and -NIT-GC2 coding sequences were amplified by PCR starting from a *Platynereis dumerilii* cDNA library and cloned into the pcDNA3.1(+) vector using the T2A self-cleaving sequence.
COS-7 cells with low expression of endogenous soluble guanylate cyclase were used as cultured cells for gene expression.
This cell line was purchased from Angio-proteomie (CAT no. cAP-0203).
The COS-7 cells were maintained at 37 °C in 35mm dishes (Nunc™ Glass Bottom Dishes) containing 3 mL of DMEM, high glucose glutamax medium (Thermo; Cat. No. 10566016) supplemented with 10% fetal bovine serum (Thermo; Cat. No. 10082147).
Upon reaching confluency of approximately 85%, we transfected the cells with the plasmid containing Green cGull-T2A-NITGC1.
Transfections were carried out with 150 ng of each plasmid and 0.3 μl of the transfection Lipofectamine 3000 Reagent (invitrogen; Cat. No. L3000001).
Two days post-transfection, we removed the culture medium and substituted it for fresh DMEM-medium.
For single-wavelength imaging experiments, cells in 35-mm dishes were washed twice and imaged in modified Ringer's buffer (140 mM NaCl, 3.5 mM KCl, 0.5mM NaH2PO4 , 0.5mM S-3 MgSO4 , 1.5 mM CaCl2 , 10 mM HEPES, 2 mM NaHCO3 and 5 mM glucose).
Dishes were mounted on a stage heated at 37 °C and imaging was performed using an inverted microscope (LSM880, Zeiss) equipped with an oil-immersion objective lens (UApo/340, 40, NA = 0.17).
Images were acquired using a xenon lamp, 460--495 nm excitation filter, 505-nm dichroic mirror and 510-- 550-nm emission filter (Zeiss).
S-Nitroso-N-acetyl-D, L-penicillamine (SNAP) was purchased from Sigma-Aldrich (St. Louis, MO, USA) .
The exposure time of the EM-CCD camera was controlled by the ZEN software (Zeiss).
Images were acquired every 15 s for 10 min and stimulation was initiated 2 min after starting image acquisition.
Imaging data analysis was performed using ImageJ (National Institutes of Health, Bethesda, MD, USA).
